Stochastic Variance-Reduced Policy Gradient

### 2018-02-07 svrpg.tex
Added definition of grad estimators (on/off)

The main limitation of plain policy gradient is the high variance of these estimators.
The na\"ive approach of increasing the batch size is not an option in RL due to the high cost of collecting samples, \ie by interacting with the environment.
For this reason, literature has focused on the introduction of baselines aiming to reduce the variance~\citep[\eg][]{Peters2008reinf,Thomas2017actionbaseline,wu2018variance}.\footnote{A baseline is a function $f(s,a)$ such that $\EVV[a \sim \pol]{f(s,a)} = 0$, for any $s\in\Sspace$.}
There has been a surge of recent interest in variance reduction techniques for gradient optimization in supervised learning.
Although these techniques have been mainly derived for finite-sum problems, we will show in Sec.~\ref{sec:alg} how they can be used in RL.
The next section has the aim to describe variance reduction techniques for finite-sum problems. In particular, we will present the SVRG algorithm that is at the core of this work.

@@

\section{Policy Gradient Estimators} \label{A:gradient_estimators}
We present a brief overview of two most widespread gradient estimators (REINFORCE~\citep{williams1992simple} and G(PO)MDP~\citep{baxter2001infinite}) both in on-policy and off-policy settings.
\subsection{On-policy}
The REINFORCE gradient estimator~\citep{williams1992simple} provides a simple, unbiased way of estimating the gradient:
\begin{align*}
\gradApp{\vtheta}{N} = \frac{1}{N}\sum_{n=1}^{N}\left(\sum_{h=0}^{H-1}\score{\vtheta}{a_h^n\vert s_h^n}\right)\left(\sum_{h=0}^{H-1}\gamma^h r_h^n - b\right),
\end{align*}
where subscripts denote the time step, superscripts denote the trajectory, $r_h$ is the reward actually collected at time $h$ and b can be any baseline provided $\EVV[a \sim \pol]{b(s,a)} = 0$, for any $s\in\Sspace$.
The G(PO)MDP gradient estimator~\cite{baxter2001infinite} is a refinement of REINFORCE which is subject to less variance \cite{zhao2011analysis} while preserving the unbiasedness:
\begin{align*}
\gradApp{\vtheta}{N} = \frac{1}{N}\sum_{n=1}^{N}\sum_{h=0}^{H-1}\left(\sum_{k=0}^{h}\score{\vtheta}{a_h^n\vert s_h^n}\right)\left(\gamma^h r_h^n - b\right).
\end{align*}

\subsection{Off-Policy}
\todopir{Make ref to Luca's thesis and describe in brief the derivations}

>>>>

The main limitation of plain policy gradient is the high variance of these estimators.
The na\"ive approach of increasing the batch size is not an option in RL due to the high cost of collecting samples, \ie by interacting with the environment.
For this reason, literature has focused on the introduction of baselines aiming to reduce the variance~\citep[\eg][]{Peters2008reinf,Thomas2017actionbaseline,wu2018variance}.\footnote{A baseline is a function $f(s,a)$ such that $\EVV[a \sim \pol]{f(s,a)} = 0$, for any $s\in\Sspace$. \todopir{check the definition of baseline}}
There has been a surge of recent interest in variance reduction techniques for gradient optimization in supervised learning.
Although these techniques have been mainly derived for finite-sum problems, we will show in Sec.~\ref{sec:alg} how they can be used in RL.
The next section has the aim to describe variance reduction techniques for finite-sum problems. In particular, we will present the SVRG algorithm that is at the core of this work.

@@

\section{Policy Gradient Estimators} \label{A:gradient_estimators}
We present a brief overview of two most widespread gradient estimators (REINFORCE~\citep{williams1992simple} and G(PO)MDP~\citep{baxter2001infinite}) both in on-policy and off-policy settings.
Let $\tau = \{\langle s_t,a_t \rangle\}_{t=0}^{H}= \{z_t\}_{t=0}^{H} = z_{0:H}$ is a $(H+1)$-steps trajectory.
Given that $\tau$ depends on the MDP $M=\{\mathcal{S},\mathcal{A}, p, r, \gamma, \rho\}$ and the actual policy $\pi$, the trajectory is
said drawn from density distribution $\Delta(\tau|\pi,M)$ defined as:
\[
        \Delta(\tau|\pi,M) = \rho(s_0) \pi(z_0) \prod_{k=1}^{H} p(s_k|z_{k-1})\pi(z_k).
\]
We can now recall the definition of policy performance 
\[
        J(\pol) = \EVV[\tau \sim \Delta(\cdot|\pol)]{\Reward(\tau)|M},
\]
where $\Reward(\tau) = \sum_{t=0}^{H}\gamma^t r(z_t)$.
The policy gradient $\nabla J(\vtheta)$ is
\begin{equation}\label{E:onpolicygradient}
        \gradJ{\vtheta} = \EVV[\tau \sim \Delta(\cdot|\pol)]{\nabla \log\Delta(\tau|\vtheta)\Reward(\tau)} = \EVV[\tau \sim \Delta(\cdot|\pol)]{\sum_{j=0}^{H} \gamma^j \mathcal{R}(z_j) \sum_{i=0}^{H} \nabla \log\pi(z_i)}.
\end{equation}

\textbf{On-policy setting.}
Consider a policy $\pol$ and let $\mathcal{D}_N = \{\tau_i\}_{i=1}^N$ be a dataset collected using policy $\pol$.
The REINFORCE gradient estimator~\citep{williams1992simple} provides a simple, unbiased way of estimating the gradient:
\begin{align*}
\gradApp{\vtheta}{N} = \frac{1}{N}\sum_{n=1}^{N}
\underbracket{
\left(\sum_{h=0}^{H}\score{\vtheta}{a_h^n\vert s_h^n}\right)\left(\sum_{h=0}^{H}\gamma^h r(s_h^n,a_h^n) - b\right)
}_{g(\tau_n|\vtheta)}
,
\end{align*}
where subscripts denote the time step, superscripts denote the trajectory, $r_h$ is the reward actually collected at time $h$ and b can be any baseline provided $\EVV[a \sim \pol]{b(s,a)} = 0$, for any $s\in\Sspace$.\todopir{check definition of baseline}
The G(PO)MDP gradient estimator~\cite{baxter2001infinite} is a refinement of REINFORCE which is subject to less variance \cite{zhao2011analysis} while preserving the unbiasedness:
\begin{align*}
\gradApp{\vtheta}{N} = \frac{1}{N}\sum_{n=1}^{N}
\underbracket{
\sum_{h=0}^{H}\left(\sum_{k=0}^{h}\score{\vtheta}{a_h^n\vert s_h^n}\right)\left(\gamma^h r(s_h^n,a_h^n) - b\right)
}_{g(\tau_n|\vtheta)}.
\end{align*}
G(PO)MDP can be seen as a more efficient implementation
of the REINFORCE algorithm. In fact the latter does not perform
an optimal credit assignment, since it ignores that the reward at time $t$ does
not depend on the action performed after time $t$. G(PO)MDP overcomes this
issue taking into account the causality of rewards in the REINFORCE definition
of policy gradient.

\textbf{Off-Policy setting.}
In off-policy setting two policies, called behavioural $\pi^B$ and target $\pi^T$, are involved.
The first is used to select actions for the interaction with the system, while the second is used to evaluate the agent performance and it is improved in each update.
Suppose now that we aim to estimate the performance of the target policy $\pi^T$ but we have samples collected using policy $\pi^B$.
We can use importance weight correction to correct the shift in the distribution and obtain an unbiased estimate of $J(\pi^T)$:
\[
        J(\pi^T) = \EVV[\tau \sim \Delta(\cdot|\pi^T)]{\mathcal{R}(\tau)} = \EVV[\tau \sim \Delta(\cdot|\pi^B)]{\omega(\tau) \mathcal{R}(\tau)}
\]
where $\omega(\tau) = \frac{\Delta(\tau,\pi^T)}{\Delta(\tau|\pi^B)} = \omega(z_{0:H}) = \prod_{w=0}^{H} \omega(z_w)$ and $\omega(z_w) = \frac{\pi^T(a_w|s_w)}{\pi^B(a_w|s_w)}$

The definition of the off-policy version of~\eqref{E:onpolicygradient} is~\citep[\eg][]{jurvcivcek2012reinforcement}
\begin{equation}\label{E:offpolicygradient}
        \nabla J(\pi^T) = \EVV[\tau \sim \Delta(\cdot|\pi^B)]{\omega(\tau) \nabla \log \Delta(\tau|\pi^T) \mathcal{R}(\tau)}
\end{equation}
For $\omega(\tau)$ being well defined the behavioural policy should have non-zero probability of selecting any action in every state \ie $\pi^B(a|s) > 0$ for any $(s,a)\in \mathcal{S} \times \mathcal{A}$.

The off-policy version of REINFORCE is easily obtained by taking the empirical average of~\eqref{E:offpolicygradient}:
\[
        \nabla J(\pi^T) = \frac{1}{N} \sum_{n=1}^{N} \omega(\tau^n)
        \left(\sum_{h=0}^{H}\score{\vtheta}{a_h^n\vert s_h^n}\right)\left(\sum_{h=0}^{H}\gamma^h r(s_h^n,a_h^n)\right).
\]

The G(PO)MDP off-policy estimator is defined as follows
\[
        \nabla J(\pi^T) = \frac{1}{N} \sum_{n=1}^{N}
        \sum_{h=0}^H \left(\sum_{k=0}^h \nabla \log\pi^T(z_k)\right) \gamma^h r(z_h^n) \omega(z_{0:h})
\]
\todopir{Add proof of Def 2 from Luca's thesis}

\todopir{Make ref to Luca's thesis and describe in brief the derivations}
